Extreme heterogeneity of influenza virus infection in single cells

### 2017-09-25 paper/paper.tex
added Sarah Teichmann citation

In the case of influenza virus infection, targeted measurements of specific proteins or RNAs have shed light on some factors that contribute to cell-to-cell heterogeneity. 
The influenza virus genome consists of eight negative-sense RNA segments, and many infected cells fail to express one more of these RNAs~\citep{Heldt:2015iz,Dou:2017cp} or their encoded proteins~\citep{Brooke:2013kb}.
In addition, activation of innate-immune responses is inherently stochastic~\citep{shalek2013single,shalek2014single,bhushal2017cell}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
However, the extent of cell-to-cell variation in these and other host and viral factors remains unclear, as does the association among them in individual infected cells.

@@

\subsection{Activation of the interferon response is rare in single infected cells.}
Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,crotta2013type}.
%In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,shalek2013single,shalek2014single,PerezCidoncha:2014jr,bhushal2017cell}.
We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.

@@

One factor that surprisingly does \emph{not} appreciably contribute to the heterogeneity in our experiments is activation of innate-immune interferon pathways.
Only one of the hundreds of virus-infected cells expresses any detectable interferon, despite the fact that a number of cells fail to express the influenza-virus interferon antagonist NS1.
It is known that interferon activation is stochastic at the level of single cells in response to both synthetic ligands~\citep{shalek2013single,shalek2014single,bhushal2017cell} and actual infection~\citep{Rand:2012kl,PerezCidoncha:2014jr,avraham2015pathogen,Killip:2017ef}.
But interferon expression is a prominent transcriptional signature of high-MOI influenza virus infection of bulk cells, including in the epithelial cell line and at the time-points used in our experiments~\citep{Geiss:2002a,sutejo2012activation}.
So it is notable how rarely single cells express interferon.
Interferon expression would surely be more common at later times or with a viral stock passaged at higher MOI, since paracrine interferon signaling~\citep{crotta2013type} and accumulation of defective viral particles enhance innate-immune detection~\citep{Tapia:2013kf,Lopez:2014en}.
However, the early events of physiological influenza infection involve just a few virions~\citep{varble2014influenza,mccrone2017evolutionary}, and so it is interesting to speculate whether rare events such as interferon activation during the first few cycles of viral replication could contribute to heterogeneity in the eventual outcome of infection.

>>>>

In the case of influenza virus infection, targeted measurements of specific proteins or RNAs have shed light on some factors that contribute to cell-to-cell heterogeneity. 
The influenza virus genome consists of eight negative-sense RNA segments, and many infected cells fail to express one more of these RNAs~\citep{Heldt:2015iz,Dou:2017cp} or their encoded proteins~\citep{Brooke:2013kb}.
In addition, activation of innate-immune responses is inherently stochastic~\citep{shalek2013single,shalek2014single,bhushal2017cell,hagai2017gene}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
However, the extent of cell-to-cell variation in these and other host and viral factors remains unclear, as does the association among them in individual infected cells.

@@

\subsection{Activation of the interferon response is rare in single infected cells.}
Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,crotta2013type}.
%In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,shalek2013single,shalek2014single,PerezCidoncha:2014jr,bhushal2017cell,hagai2017gene}.
We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.

@@

One factor that surprisingly does \emph{not} appreciably contribute to the heterogeneity in our experiments is activation of innate-immune interferon pathways.
Only one of the hundreds of virus-infected cells expresses any detectable interferon, despite the fact that a number of cells fail to express the influenza-virus interferon antagonist NS1.
It is known that interferon activation is stochastic at the level of single cells in response to both synthetic ligands~\citep{shalek2013single,shalek2014single,bhushal2017cell,hagai2017gene} and actual infection~\citep{Rand:2012kl,PerezCidoncha:2014jr,avraham2015pathogen,Killip:2017ef}.
But interferon expression is a prominent transcriptional signature of high-MOI influenza virus infection of bulk cells, including in the epithelial cell line and at the time-points used in our experiments~\citep{Geiss:2002a,sutejo2012activation}.
So it is notable how rarely single cells express interferon.
Interferon expression would surely be more common at later times or with a viral stock passaged at higher MOI, since paracrine interferon signaling~\citep{crotta2013type} and accumulation of defective viral particles enhance innate-immune detection~\citep{Tapia:2013kf,Lopez:2014en}.
However, the early events of physiological influenza infection involve just a few virions~\citep{varble2014influenza,mccrone2017evolutionary}, and so it is interesting to speculate whether rare events such as interferon activation during the first few cycles of viral replication could contribute to heterogeneity in the eventual outcome of infection.
